---
title: Relationship between enhancer methylation, promoter methylation, and expression
  in ccRCC
author:
- name: Hamed S. Najafabadi
#date: "`r format(Sys.Date(), "%Y-%B-%d")`"
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
editor_options: 
  chunk_output_type: inline
---

# Preparation
Load the files and libraries:
```{r}

library(ggplot2)
library(ggforce)
library(gridExtra)
library(scales)
library(randomForest)
library(IHW)
library(pROC)
library(fields)
library(matlab)
library(stringr)
library(MASS)


# Load the environment from the previous step
#load("01.gain_loss_promoterss.table.RData")
#exp_table_p <- table
#table <- exp_table_p
```

# How does gain/loss of enhancers determine gene expression changes in tumor vs. normal?
Here, we will use the gain/loss score of enhancers together with machine learning to explore the relationship with gene expression changes. The predictors are the gain/loss score and the enhancer distance to the gene. The target is the status of the gene (significant up-regulation, significant down-regulation, or no change)

```{r}
# create a data frame of the covariates, including the target for prediction
covariates <- data.frame(
  target = as.factor( (table$padj_patients_T_vs_N < 0.01)*1 * sign(table$log2FoldChange_patients_T_vs_N) ),
  enhancer_score = table$gain_model_score,
  distance = table$distance_to_the_gene_TSS_from_centre_of_enhancer )

# train the model using default parameters of the random forest function
TN_gene_expression_model <- randomForest( target ~ ., data=covariates, na.action = na.omit )
# predict gain/loss using this model
predicted <- predict(TN_gene_expression_model, covariates, type="prob")
colnames(predicted)[1:3] <- c("expression_down_model_score","expression_nochange_model_score","expression_up_model_score")
table <- cbind( table, predicted )
table$predicted_upregulated <- table$expression_up_model_score > 0.7
table$predicted_downregulated <- table$expression_down_model_score > 0.7
table$predicted_nochange <- table$expression_nochange_model_score > 0.7

summary(table$predicted_upregulated)
summary(table$predicted_downregulated)
summary(table$predicted_nochange)

```

Let's perform cross-validation to examine the performance (leave-one-chromosome-out):
```{r}
cv.result <- cv.rf(target ~ .,covariates, id=table$chr) # use only the non-redundant enhancers (the ones marked with TRUE), and use "chr" as the id for cross-validation

#pdf(file="C:/Users/pubud/Dropbox/thesis/graphs/supfig3/supfig3a_promoters__genes_roc.pdf", height=5,width=8)
par(mfrow=c(1,2))
proc <- roc( (cv.result[,1]=="1"), cv.result[,ncol(cv.result)] ) # the first column of cv.result is the target, and the last column is the predicted probability of being up-regulated
cutoff <- min( which(proc$threshold > 0.7) )
print(paste0("At 0.9 cutoff, the sensitivity and specificity for predicting up-regulated genes is ", round(proc$sensitivities[cutoff],2), " and ", round(proc$specificities[cutoff],2), ", respectively."))
plot(proc,main=str_wrap("Predicting up-regulated genes",20))
#lines( proc$specificities, proc$thresholds, lty=2 )

proc <- roc( (cv.result[,1]=="-1"), cv.result[,ncol(cv.result)-2] ) # the first column of cv.result is the target, and the last column is the predicted probability of being down-regulated
cutoff <- min( which(proc$threshold > 0.7) )
print(paste0("At 0.7 cutoff, the sensitivity and specificity for predicting down-regulated genes is ", round(proc$sensitivities[cutoff],2), " and ", round(proc$specificities[cutoff],2), ", respectively."))
plot(proc,main=str_wrap("Predicting down-regulated genes",20))
#lines( proc$specificities, proc$thresholds, lty=2 )

#dev.off()

```
```{r}
proc

 roc.area( (cv.result[,1]=="1"), cv.result[,ncol(cv.result)] ) 
 
 roc.area( (cv.result[,1]=="-1"), cv.result[,ncol(cv.result)-2] )
```

The classifier seems to work really well! It is important to note that in reality we are not predicting the up-/down-regulated genes, but instead we are predicted whether an enhancer is associated with an up-/down-regulated gene. The reason these are two different tasks is that each gene can be associated with multiple enhancers, and our unit of prediction is the enhancer itself.

## What are the inner workings of this classifier?

What is the relationship between enhancer state and gene expression?
```{r}
resolution <- 30
dist_range <- 20000
positive <- covariates[ which(predicted[,3]>0.7) , ]
negative <- covariates[ which(predicted[,3]<0.7) , ]
lims=c(-dist_range,dist_range,-0.05,1.05)
f1 <- kde2d(positive$distance,positive$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
f2 <- kde2d(negative$distance,negative$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
z <- f1$z-f2$z
#z <- log10( (f1$z+mean(f1$z)*10)/(f2$z+mean(f2$z)*10) )

z_range <- max(abs(range(z)))
z[ z < -z_range ] <- -z_range
z[ z > z_range ] <- z_range
image(f2$x,f2$y,z,zlim=c(-z_range,z_range),col=colorRampPalette(c("blue", "white","red"))(100),
      xlab="Distance from TSS", ylab="Promoter gain score", main="Predicted up-regulation")


resolution <- 30
dist_range <- 20000
positive <- covariates[ which(predicted[,1]>0.7) , ]
negative <- covariates[ which(predicted[,1]<0.7) , ]
f1 <- kde2d(positive$distance,positive$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
f2 <- kde2d(negative$distance,negative$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
z <- f1$z-f2$z

z_range <- max(abs(range(z)))
z[ z < -z_range ] <- -z_range
z[ z > z_range ] <- z_range
image(f2$x,f2$y,z,zlim=c(-z_range,z_range),col=colorRampPalette(c("blue", "white","red"))(100),
      xlab="Distance from TSS", ylab="Promoter gain score", main="Predicted down-regulation")


resolution <- 30
dist_range <- 20000
positive <- covariates[ which(predicted[,2]>0.7) , ]
negative <- covariates[ which(predicted[,2]<0.7) , ]
f1 <- kde2d(positive$distance,positive$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
f2 <- kde2d(negative$distance,negative$enhancer_score,lims=lims,n=500,h=c(dist_range*2/resolution,1/resolution))
z <- f1$z-f2$z

z_range <- max(abs(range(z)))
z[ z < -z_range ] <- -z_range
z[ z > z_range ] <- z_range
image(f2$x,f2$y,z,zlim=c(-z_range,z_range),col=colorRampPalette(c("blue", "white","red"))(100),
      xlab="Distance from TSS", ylab="Promoter gain score", main="Predicted no-change")

```


# What is the relationship between epigenetic modulation of gene expression in ccRCC and VHL-mediated changes in gene-expression?

Let's look at the enhancers that are predicted to be associated with up- or down-regulated genes, and see whether they are overall associated with VHL-mediated up-/down-regulated genes
```{r}

VHL_H <- as.factor( (table$padj_VHL_vs_EV_H < 0.01)*1 * sign(table$log2FoldChange_VHL_vs_EV_H) )
VHL_N <- as.factor( (table$padj_VHL_vs_EV_N < 0.01)*1 * sign(table$log2FoldChange_VHL_vs_EV_N) )

hypo.results <- enrichment2D(
  VHL_H,
  VHL_N,
  table$predicted_upregulated | table$predicted_downregulated,
  table$predicted_upregulated,
  str_wrap("VHL status, hypoxia",width=30),
  str_wrap("VHL status, normaxia",width=30),
  logOddsLimit = 0.1 )

hypo.results$plot + ggtitle("Enrichment of enhancer-upregulated genes")


```

This graph clearly shows that genes that are down-regulated by VHL both in hypoxia and normaxia are most highly enriched for enhancers that are gained in tumor (more accurately, for genes that are predicted to be up-regulated in tumor because of a nearby gained enhancer). These are "hypoxia-independent VHL-repressed genes". Interestingly, genes that are down-regulated by VHL in normaxia, but not in hypoxia are also enriched, suggesting that hypoxia-dependent VHL repression also plays a role in shaping tumor expression. However, it is important to note that only a fraction of gained enhancers are in these two categories, and the large majority of gained enhancers are not affected by VHL:

```{r}
cat( paste0( "The fraction of enhancer-mediated gene up-regulation that is recaptured by VHL is ", round( sum(VHL_N=="-1" & table$predicted_upregulated, na.rm=T)/sum(table$predicted_upregulated, na.rm=T), digits = 2 ), "\n" ) )

```


```{r}

lims=c(-1,1,-1,1)
filter <- !is.na(table$log2FoldChange_VHL_vs_EV_H) & !is.na(table$log2FoldChange_VHL_vs_EV_N)
positive <- table[ which(table$predicted_upregulated & filter), ]
negative <- table[ which(table$predicted_downregulated & filter), ]
f1 <- kde2d(positive$log2FoldChange_VHL_vs_EV_H,positive$log2FoldChange_VHL_vs_EV_N,lims=lims,n=500,h=0.6)
f2 <- kde2d(negative$log2FoldChange_VHL_vs_EV_H,negative$log2FoldChange_VHL_vs_EV_N,lims=lims,n=500,h=0.6)
z <- f1$z-f2$z

z_range <- max(abs(range(z)))
z[ z < -z_range ] <- -z_range
z[ z > z_range ] <- z_range
image(f2$x,f2$y,z,zlim=c(-z_range,z_range),col=colorRampPalette(c("dark blue", "blue", "white","red", "dark red"))(50))
#abline(h=0,lt=2)
#abline(v=0,lt=2)
contour(f2$x, f2$x, z, col = "black", add = TRUE, method = "edge",
        vfont = c("sans serif", "plain"), labcex=1)

```

Write the list of these genes and their associated enhancers:
```{r}


write.table( table[ which(
    VHL_N=="-1" &
    VHL_H=="-1" &
    table$predicted_upregulated &
    table$predicted_gain &
    table$padj_patients_T_vs_N < 0.01 &
    table$log2FoldChange_patients_T_vs_N > 0),],
    file="02.VHL_suppressed.hypoxia_independent.promoter_mediated_tumor_upregulation.csv",
    quote=T, row.names = F, sep="," )

```



Gained and lost enhancers and gene expression correlation

```{r}
predicted_element <- fread("02.promoter_vs_expression.table.csv")

predicted_element <- subset(predicted_element, !is.na(gain_model_score ) & !is.na(log2FoldChange_patients_T_vs_N) & padj_patients_T_vs_N < 1)

x = predicted_element$gain_model_score
smoothscattergradient( x=x, y=predicted_element$log2FoldChange_patients_T_vs_N, file.name=NULL , xlabs="PREs Model Score", ylabs="Log2 FC Patients", width=5, height=5, lab.x=0.5, lab.y=10, bin.size=75,bin.width1=0.1,bin.width2=0.1 , my.breaks=c(1,10, 100), up.lab.x=4,up.lab.y=-4.5, down.lab.x=-2.5,down.lab.y=7.5, returnfraction=F)
```

Gained and lost enhancers and gene expression correlation

```{r}
predicted_element <- fread("02.promoter_vs_expression.table.csv")

predicted_element <- subset(predicted_element, !is.na(gain_model_score ) & !is.na(log2FoldChange_patients_T_vs_N) )

#sort by abstract distance
#predicted_element <- predicted_element[order(predicted_element$gain_model_score),]
#remove duplicates by gene name
#predicted_element <- predicted_element[!duplicated(predicted_element$Gene.name),]



x = predicted_element$gain_model_score
y= predicted_element$log2FoldChange_patients_T_vs_N
smoothscattergradient( x=x, y=y, file.name=NULL , xlabs="Promoter Model Score", ylabs="Log2 FC Patients", width=5, height=5, lab.x=0.5, lab.y=10, bin.size=80,bin.width1=0.1,bin.width2=0.1 , my.breaks=c(1,10, 100), up.lab.x=4,up.lab.y=-4.5, down.lab.x=-2.5,down.lab.y=7.5, returnfraction=F)


y2 <- transform( y, y = sample(y) )


smoothscattergradient( x=x, y=y2$y, file.name=NULL , xlabs="Promoter Model Score", ylabs="Log2 FC Patients", width=5, height=5, lab.x=0.5, lab.y=10, bin.size=80,bin.width1=0.1,bin.width2=0.1 , my.breaks=c(1,10, 100), up.lab.x=4,up.lab.y=-4.5, down.lab.x=-2.5,down.lab.y=7.5, returnfraction=F)
```


```{r}

lims=c(0,1,-10,10)


f1 <- kde2d(x,y, lims=lims,n=500,h=3.9)
f2 <- kde2d(x,y2$y, lims=lims,n=500,h=3.9)
z <- f1$z-f2$z
z_range <- max(abs(range(z)))
z[ z < -z_range ] <- -z_range
z[ z > z_range ] <- z_range
pdf(file="C:/Users/pubud/Dropbox/thesis/graphs/fig 4/fig4_promoter_score_and_log2fc.pdf", height=5,width=7)
image(f2$x,f2$y,z,zlim=c(-z_range,z_range),col=colorRampPalette(c("blue", "white","red"))(255))
#abline(h=0,lt=2)
#abline(v=0,lt=2)
#contour(f2$x, f2$y, z, col = "black", add = TRUE, method = "edge",
 #       vfont = c("sans serif", "plain"), labcex=1.5)
dev.off()
```



```{r}
filled.contour(f2$x,f2$y,z,zlim=c(-z_range,z_range),color.palette=colorRampPalette(c("blue","white","red" )),
               levels = pretty(c(-z_range,z_range), 255), cex.axis = 2.6, cex.lab = 2.8, cex.main=4)
```

# Write the updated table and the environment variables, and show the session info
```{r}
write.table( table, "02.promoter_vs_expression.table.csv", sep=",", quote=T, row.names=F )
save(list=ls(),file="02.promoter_vs_expression.table.RData")

sessionInfo()
```
